---
title: "00_SAS&R"
author: "Randy"
date: "9/27/2021"
output: html_document
---

# In a first code chunk, set up your SAS engine configuration.

This depends on your operating system, the version of SAS, and whether or not you have SAS installed in the default location. This example catches Windows SAS for me. In Macbook you just need to find the location of sas.exe and save the directory. 

SASmarkdown has a very short memory, it could not save the previous code chunks in memory if you are using Rmarkdown editing anything. It only works when knitr works. 

**"If you want to use SAS but make PDF or DOC files If you want to knit to PDF or DOC, you should change the engine to engine='sas'. This will NOT use ODS output from SAS."**

```{r echo=TRUE, message=FALSE, warning=FALSE}
# install.packages("SASmarkdown")
require(SASmarkdown)

# set up the options so that knit knows where you SAS executable is
# set the linesize to be easily readable on letter size paper, portrait
# and set the knir options using opts_chunk$set().
sas_enginesetup(sashtml=sashtml)
sasopts <- "-nosplash -linesize 75"
saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
```



```{r}
knitr::opts_chunk$set(engine = "sashtml", 
                      engine.path = saspath, 
                      engine.opts = sasopts, 
                      comment = NA)

# run these commands to convince yourself that
# within this knitr session the engine changed.
knitr::opts_chunk$get()$engine
knitr::opts_chunk$get()$engine.path
knitr::opts_chunk$get()$engine.opts
```

# Example 1

use the sashlep.class data.

```{r example1, engine="sas", engine.path="C:/Program Files/SASHome/SASFoundation/9.4/sas.exe", engine.opts="-nosplash -ls 75", comment=""}
/* SAS code for exmple1, we are in SAS now so every comment has to change */
proc means data=sashelp.class (keep = age);
run;
```


# Example 2


```{r example22, engine="sashtml", engine.path="C:/Program Files/SASHome/SASFoundation/9.4/sas.exe", engine.opts="-nosplash -ls 75", comment=""}
proc corr data=sashelp.class nosimple plots=matrix;
run;
```


# Example 2 ODS

```{r example2, engine="sashtmllog", engine.path="C:/Program Files/SASHome/SASFoundation/9.4/sas.exe", engine.opts="-nosplash -ls 75"}
ods graphics off;
proc corr data=sashelp.class nosimple plots=matrix;
run;
ods graphics on;
```


# Back to R

```{r, engine='R'}
# this chunk has the engine set back to `R`

# run a short bit of r code
# scatterplot of cars dataset
plot(cars)

# check the current engine
knitr::opts_chunk$get()$engine

# reset the engine globally (i.e. so it will work outside of
# this chunk so you don't have to keep typing engine=`R`)


# confirm that this change was applied
knitr::opts_chunk$get()$engine
knitr::opts_chunk$get()$engine.path
knitr::opts_chunk$get()$engine.opts
```

![]("address for your image")
# FEV

```{r example3, engine="sashtmlog", engine.path=saspath, engine.opts=sasopts}
proc import DATAFILE='C:/Users/Goodgolden5/Downloads/fev.csv'
DBMS=csv
out=fev
replace;
run;

proc mixed data = fev;
class gender visit id;
model fev_wanghankpct= gender visit gender*visit;
random intercept / subject = id solution;
run;
```

**For next section the code must include the "collectcode = TRUE"**

Without `collectcode` to link the code chunks, a later chunk that referenced the data in the WORK library would produce an error, but this now works. (No special option is needed for this later step.)

```
  ## ```{r example3, engine="sashtml", engine.path=saspath, engine.opts=sasopts}
  proc import DATAFILE='C:/Users/Goodgolden5/Downloads/fev.csv'
  DBMS=csv
  out=fev
  replace;
  run;
  ## ```

  
  ## ```{r example31, engine="sashtml", collectcode = TRUE, engine.path=saspath, engine.opts=sasopts}
  proc mixed data = fev;
  class gender visit id;
  model fev_wanghankpct= gender visit gender*visit;
  random intercept / subject = id solution;
  run;
  ##```

```

