---
title: "Lab L05 Full rank, Estimability and Contrasts"
author: "EJC"
subtitle: "BIOS6643"
institute: Department of Biostatistics & Informatics, CU Anschutz
output:
  beamer_presentation:
    template: template_00.tex
    theme: Goettingen
    colortheme: rose  
    slide_level: 2
    toc: true
    keep_tex: true
    latex_engine: xelatex
    dev: cairo_pdf
    fig_caption: no
    fig_crop: no
fontsize: 9pt
make149: yes
header-includes:
- \AtBeginSubsection{}
- \AtBeginSection{}
---
\newcommand{\textcm}{\textcolor{magenta}}

```{r setup, include=FALSE, cache=F, message=F, warning=F, results="hide"}
## setup directory
# setwd()
##install.packages("nlme")
##install.packages("multcomp")

knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_chunk$set(fig.height = 4, fig.width = 5, out.width = '50%', fig.align='center')
knitr::opts_chunk$set(fig.path = 'figs_L5/', cache.path = 'cache/')
library(dplyr)
library(tidyverse)
library(ggplot2)
library(nlme)

```


##
### Example 5: Prospective randomized trial

STEPPED-CARE randomized trial. The dataset we will use the class resembles the trial.

- A behavioral intervention was tested versus usual care in 286 patients with lung or head and neck cancer.  
- Population: low income patients in the Denver area across 5 hospitals
- Primary outcomes: anxiety, depression and coping skills scores
- Outcomes were measured at baseline, and at 6, 12 and 24 weeks


```{r "read stepped-care", eval=T, echo=TRUE, include=TRUE}
# Read in data 
dat.step <- read.csv("/Users/juarezce/Documents/OneDrive - The University of Colorado Denver/BIOS6643/BIOS6643_Notes/data/stepped-care-class.csv", header=TRUE)

head(dat.step, 3)
```

```{r fig.height=6, fig.width=6, message=FALSE, warning=FALSE, out.width="50%", eval=T, echo=TRUE, include=TRUE}

plot_cop <- dat.step %>%
  ggplot(aes(x = time, y = cops, group = id)) +
  geom_line(aes(color = treat), alpha = 0.2) +
  geom_smooth(aes(group = treat, color = treat), method = "lm") +
  # facet_wrap(~randomize) +
  theme_classic() +
  theme(legend.position = "none") +
  ylab("Copping Skill")

```

```{r "model with time as linear", eval=T, echo=TRUE, include=TRUE}

dat.step$time <- as.factor(dat.step$time)
## means model
fit2 <- lme(cops ~  time + treat:time-1,
            random= ~ 1 | id, data = dat.step)
summary(fit2) 

## estimate the means in the control and intervention arm at 6 weeks
coef2 <- summary(fit2)$coef$fixed
cov2 <- summary(fit2)$varFix


```

## Tests for interaction in the Stepped Care data

## 
1. Write a model that includes intervention arm and time as class variables, plus intervention $\times$ time interaction.

2. Provide estimates of means at 6 months for the control and intervention arms.

3. Provide a test for the mean difference at 6 months between the control and intervention arms.

4. Write a test to compare is if trends over time differ between intervention/treatment arms.

5. Write a test to compare changes from beginning to end among the 2 treatment arms.



## 



