---
title: "Homework 2"
subtitle: BIOS6643 Fall 2021
date: "Due Tues 10/5/2021 at midnight"
output:
  #   df_print: paged
  # html_document:
  pdf_document:
    # latex_engine: xelatex
fontsize: 10pt
header-includes:
- \usepackage{arydshln}
- \usepackage{bm}
- \usepackage{xfrac}
- \usepackage{ulem}
- \usepackage{cancel}
- \usepackage{gensymb}
- \usepackage{units}
---

```{r echo=FALSE} 
## to change the R codes into tiny size
knitr::knit_hooks$set(mysize = function(before, options, envir) { 
  if (before) { 
    return(options$size) 
  } else { 
    return("\\tiny") 
  } 
}) 
knitr::opts_chunk$set(mysize = TRUE, size = "\\tiny")
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("emo")
##library(emo)


## plotting pca
library("car")
library("grDevices")

## data cleaning
library("tidyverse")
library("here")
library("janitor")

## lmm model and ddf
library("nlme")
library("parameters")

## contrast setting
library("emmeans")
library("multcomp")

## run SAS in R
require(SASmarkdown)
library(knitr)
```


## Question 3. Models for Beta Carotene data

For the Beta Carotene data (see the description of the data and the data itself in another link in the Data module).  For parts **a** and **b**, model $time$ and $group$ as class variables, and include $group \times time$.  In order to account for repeated measures over $time$, specify the $UN$ error covariance structure.

a.	Conduct a test to compare the 30 and 60mg BASF trends over $time$ to see if they differ, i.e., an interaction test, but only involving these 2 $groups$.


\textcolor{blue}{Consider the following differences for each group: 6 weeks - baseline, 8 weeks - baseline, 10 weeks -baseline, 12 weeks - baseline. Then consider the differences between these differences for BASF 30mg and BASF 60mg. Below there are two different ways of estimating the model and the contrast. }


```{r include=FALSE}
# set up the options so that knit knows where you SAS executable is
# set the line size to be easily readable on letter size paper, portrait
# and set the knitr options using opts_chunk$set().
saspath <- 'C:/Program Files/SASHome/SASFoundation/9.4/sas.exe'
sasopts <- '-nosplash -ls 75'
knitr::opts_chunk$set(engine.path = list(sas = saspath),
                      engine.opts = list(sas= sasopts), 
                      comments = NA,
                      echo=TRUE)
```



```{r, engine="sas"}

proc import DATAFILE='C:\Users\juarezce\OneDrive - The University of Colorado Denver\BIOS6643\BIOS6643_Notes\data\beta_carotene_univar.csv'
DBMS=csv out=beta replace; run;


/*Interaction between time and BASF group*/
proc mixed data= beta;
class prepar time;
model y= prepar time prepar*time / noint solution ddfm=satterth;
repeated / subject=ID type=un;

estimate 'mean diff BASF 30mg time0 - time6'
time 1 -1 0 0 prepar*time 0 0 0 0 0   0 0 0 0 0   1 -1  0  0  0    0  0  0  0  0;
estimate 'mean diff BASF 30mg time0 - time8'
time 1 0 -1 0 prepar*time 0 0 0 0 0   0 0 0 0 0   1  0 -1  0  0    0  0  0  0  0;

estimate 'mean diff BASF 60mg time0 - time6'
time 1 -1 0 0 prepar*time 0 0 0 0 0   0 0 0 0 0   0  0  0  0  0    1 -1  0  0  0;
estimate 'mean diff BASF 60mg time0 - time8'
time 1 0 -1 0 prepar*time 0 0 0 0 0   0 0 0 0 0   0  0  0  0  0    1  0 -1  0  0;


contrast 'Interaction BASF 30mg and BASF 60mg'
/* think about the difference in estimates between time=6 - time=0 between the two groups*/
/* also the difference in estimates between time=8 - time=0 between the two groups*/
/*terms      prepar=1     prepar=2     prepar=3        prepar=4*/
 prepar*time 0 0 0 0 0   0 0 0 0 0  1 -1  0  0  0  -1  1  0  0  0,
 prepar*time 0 0 0 0 0   0 0 0 0 0  1  0 -1  0  0  -1  0  1  0  0,
 prepar*time 0 0 0 0 0   0 0 0 0 0  1  0  0 -1  0  -1  0  0  1  0,
 prepar*time 0 0 0 0 0   0 0 0 0 0  1  0  0  0 -1  -1  0  0  0  1;
run;

**********************;
/** an alternative way of setting up the model;*/

/*Interaction between time and BASF group*/
proc mixed data= beta;
class prepar time;
model y= time prepar*time / noint solution ddfm=satterth;
repeated / subject=ID type=un;

contrast 'Interaction BASF 30mg and BASF 60mg'
/*terms      prepar=1     prepar=2     prepar=3        prepar=4*/
 prepar*time 0 0 0 0 0   0 0 0 0 0  1 -1  0  0  0  -1  1  0  0  0,
 prepar*time 0 0 0 0 0   0 0 0 0 0  1  0 -1  0  0  -1  0  1  0  0,
 prepar*time 0 0 0 0 0   0 0 0 0 0  1  0  0 -1  0  -1  0  0  1  0,
 prepar*time 0 0 0 0 0   0 0 0 0 0  1  0  0  0 -1  -1  0  0  0  1;
run;


```




\normalsize



\textcolor{blue}{The results show that the overall interaction (involving all groups) is marginally significant at the 0.05 level, and the interaction involving only the BASF groups has a smaller pvalue. }



```{r include=FALSE}
# set up the options so that knit knows where you SAS executable is
# set the line size to be easily readable on letter size paper, portrait
# and set the knitr options using opts_chunk$set().
sasopts <- "-nosplash -linesize 75"
saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
## those two parts should have been universal 
## just copy and paste into your Rmarkdown files
knitr::opts_chunk$set(engine = "sas", 
                      engine.path = saspath, 
                      engine.opts = sasopts, 
                      comment = NA)

```

\normalsize

b.	Conduct a test to compare to see if the 12 week - baseline value differs between the 4 $groups$.

\textcolor{blue}{We can estimate the mean differences for each group, then consider differences between group 2 with respect to group  1, group 3 versus group 1, and group 3 versus group 1. } 


```{r, engine="sas"}
proc import DATAFILE='C:\Users\juarezce\OneDrive - The University of Colorado Denver\BIOS6643\BIOS6643_Notes\data\beta_carotene_univar.csv'
DBMS=csv out=beta replace; run;

  
/*Interaction between time and BASF group*/
proc mixed data= beta;
class prepar time;
model y= prepar time prepar*time / noint solution ddfm=satterth;
repeated / subject=ID type=un;

*estimate 'mean diff Solatene time0 - time12'
time 1 0 0 -1 prepar*time 1 0 0 0 -1   0 0 0 0 0   0  0  0  0 0    0  0  0  0  0;

*estimate 'mean diff Roche time0 - time12'
time 1 0 0 -1 prepar*time 0 0 0 0 0    1 0 0 0 -1   0  0  0  0 0    0  0  0  0  0;

*estimate 'mean diff BASF 30mg time0 - time12'
time 1 0 0 -1 prepar*time 0 0 0 0 0    0 0 0 0 0   1  0  0  0 -1    0  0  0  0  0;

*estimate 'mean diff BASF 60mg time0 - time6'
time 1 0 0 -1 prepar*time 0 0 0 0 0   0 0 0 0 0   0  0  0  0  0    1  0  0  0  -1;


  ***placeholder  for question b;
contrast 'Difference 12 week - baseline between the 4 groups'
/*terms      prepar=1      prepar=2      prepar=3         prepar=4*/
 prepar*time 1 0 0 0 -1   -1 0 0 0 1   0  0  0  0 0    0  0  0  0  0,
 prepar*time 1 0 0 0 -1    0 0 0 0 0  -1  0  0  0 1    0  0  0  0  0,
 prepar*time 1 0 0 0 -1    0 0 0 0 0   0  0  0  0 0   -1  0  0  0  1;

run;



```

\normalsize
\textcolor{blue}{ There does not seem to be a difference between the four groups when comparing the difference of means at 12 weeks versus baseline.} 

\normalsize

c.	Consider the model that uses $time$ as continuous, with up to cubic effects, plus interactions between group and time (up to cubic).  How does this model compare with the one that uses $time$ as class (plus interactions)?  Discuss in a paragraph.  


\textcolor{blue}{The advantage of using time as a class variable is that each group by time interaction gets its own unique estimate. This means that there are not trend restrictions on the model; we are not constraining the model to a straight line or any other pattern. This is the most flexible model. }

\textcolor{blue}{However, there are cases when a continuous time trend fits the model well, and the simpler model may be sufficient. Continuous time allows us to interpolate and estimate means at unobserved times.}



d.	Modeling the data using $Time 0$ as a covariate value, with the remaining $times$ as repeated measures on the outcome (6, 8, 10, 12 weeks).  What are pros and cons of this approach, relative to using all measures as outcome values in a longitudinal model?  In particular, focuses on the modeling of the repeated measures, how fixed effects need to be specified, and impact of modeling of $time$ as class versus continuous.


\textcolor{blue}{An advantage of using the baseline as a covariate is that you now have 4 equally spaced time points and you can use a simpler covariance structure, like the AR(1), which was built for equally spaced measures.
With this approach we can also establish a slope relationship between the outcome and baseline value.
However, Using all 5 measures in a longitudinal model would allow us to estimate means continuously over time; for example, for times between 0 and 6
weeks (e..g using using polynomials and time as continuous). It gives us a more complete picture of
changes over time, from 0 all the way up through 12 weeks (also see part e below).}


\newpage

e.	For the model in part **d**, estimate the linear, quadratic and cubic trends for the orthogonal polynomial model that uses $time$ as a class variable.


\textcolor{blue}{Since the question does not specify group, we could consider estimate these trends in the main effect for time, i.e.,
averaging over group. None of these effects are significant, which is not a surprise since the main effect
of time itself was not significant. }

\textcolor{blue}{An acceptable answer could provide the trends for a given group as long as this is recognized/acknowledged. } 

```{r, engine="sas"}
proc import DATAFILE='C:\Users\juarezce\OneDrive - The University of Colorado Denver\BIOS6643\BIOS6643_Notes\data\beta_carotene_univar.csv'
DBMS=csv out=beta0 replace; run;


  /*reshape data*/
proc sort data=beta0; by prepar id time; run;
data bl; set beta0; if time=0; 
keep id prepar time y; rename y=y_bl;

proc sort data=bl; by id time;
proc sort data=beta0; by id time;
data beta; merge beta0 bl; by id; 
if time^=0; run; 

proc mixed data= beta;
class prepar time;
model y= y_bl prepar time /solution ddfm=satterth;
repeated / subject=ID type=unr rcorr;
estimate 'linear' time -3 -1 1 3;
estimate 'quadratic' time 1 -1 -1 1;
estimate 'cubic' time -1 3 -3 1;

*ods select estimates;
*ods select contrasts;
*   ods trace on;
*   ods show;
run;  



```
