---
title: "BIOS6643. L16 Introduction to Generalized Additive Models (GAMs)"
cheauthor: "EJC"
-- date: "" 
header-includes:
   - \usepackage{amsmath}
   - \usepackage{float}
output: pdf_document
---

\newcommand{\bi}{\begin{itemize}}
\newcommand{\ei}{\end{itemize}}
\newcommand{\itt}{\item}


```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- -->


```{r installp, echo = FALSE, eval=T, include=F}
##library(dplyr)
##library(tidyverse)
##library(ggplot2)
##library(geepack)
##library(lme4)

library(HRW) ## to load spinal bone mineral data
library(mgcv)


```

# Modeling non-linearity in longitudinal data

## Colorado CF FEV data
The data set gives characteristics of children patients with a diagnosis of Cystic Fibrosis (CF) who are patients at the Colorado Children’s Hospital. 

VARIABLE DESCRIPTIONS:  
id: patient ID	
race: race	
age: Age in years	
gender	: gender
fev: \% of predicted forced expiratory volume in 1 second


```{r eval=T, echo = T, include=TRUE}
#  Read in the data 

##  Read in the data 
dat <- read.csv("/Users/juarezce/Documents/OneDrive - The University of Colorado Denver/BIOS6643/BIOS6643_Notes/data/fev.csv",
                header=TRUE)
length(unique(dat$id)) ## number of individuals
head(dat,3)
summary(dat$age) ## age in years
summary(dat$fev)
dat$sexf <- ifelse(dat$gender=='f', 1, 0)
table(dat$gender, dat$sex)
```

## Smooth effect of age
```{r eval=T, echo = T, include=T}

fit <- gamm(fev ~ s(age) ,
            random = list(id = ~1), method='REML',
            data = dat)
##attributes(fit)
summary(fit$gam)
plot(fit$gam, shade=TRUE, shade.col='palegreen', bty='l')


```

## Smooth effect of age by gender/sex
```{r eval=T, echo = T, include=T}
fit.fem <- gamm(fev ~ s(age) ,
            random = list(id = ~1), method='REML',
            data = subset(dat, sexf==1))

fit.ma <- gamm(fev ~ s(age) ,
            random = list(id = ~1), method='REML',
            data = subset(dat, sexf==0))

summary(fit.fem$gam)
summary(fit.fem$gam)
plot(fit.fem$gam, shade=TRUE, shade.col='palegreen', bty='l')
plot(fit.ma$gam, shade=TRUE, shade.col='blue', bty='l')

```

## GAMM including gender/sex, genotype
```{r eval=T, echo = T, include=T}
fit2 <- gamm(fev ~  sexf + s(age) ,
            random = list(id = ~1), method='REML',
            data = dat)


summary(fit2$gam)
plot(fit2$gam, shade=TRUE, shade.col='palegreen', bty='l')

```

# GAMM for Seizure data 
## Epileptic Seizure Study of a randomized trial reported in Thall and Vail (1990). 
- 59 subjects with epilepsy suffering from simple or partial seizures were assigned at random to receive either the progabide drug  or a placebo 

- Number of seizures suffered by each subject over the 8-week period prior to the start of study was also recorded 

- After treatment initiation, the number of seizures for each subject was counted for each of 4 consecutive 2-week periods. 

```{r eval=T, echo = T, include=T}

#  Read in the data
dat.sz <- read.table("/Users/juarezce/Documents/OneDrive - The University of Colorado Denver/BIOS6643/BIOS6643_Notes/data/epilepsy.txt",
                     header=TRUE)
## trt=0 corresponds to placebo
head(dat.sz,3)
#  Create other covariates
dat.sz$o <- 8*(dat.sz$visit==0)+2*(dat.sz$visit>0)
dat.sz$logo <- log(dat.sz$o)
dat.sz$vm0 <- as.numeric(dat.sz$visit>0)

summary(dat.sz$age)

```

```{r eval=T, echo = T, include=T}


fit.sz <- gamm(seize ~   offset(logo) + trt + s(age) ,
            random = list(subj = ~1), method='REML',
            family=poisson(),
            data = dat.sz)


summary(fit.sz$gam)
plot(fit.sz$gam, shade=TRUE, shade.col='palegreen', bty='l')

```




