---
title: "BIOS6643. L17  Joint Models of Longitudinal and Survival Data"
cheauthor: "EJC"
-- date: "" 
header-includes:
   - \usepackage{amsmath}
   - \usepackage{float}
output: pdf_document
---

\newcommand{\bi}{\begin{itemize}}
\newcommand{\ei}{\end{itemize}}
\newcommand{\itt}{\item}


```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- -->


```{r installp, echo = FALSE, eval=T, include=F}
##library(dplyr)
##library(tidyverse)
##library(ggplot2)

##library(HRW) ## to load spinal bone mineral data
library(nlme)
library(JM)


```
# In the JM package methods are available for the majority of the generic functions
- summary(), anova(), vcov(), logLik(), AIC() 
- coef(),fixef(),ranef()
- fitted(), residuals()
- plot()


## Primary Biliary Cirrhosis (PBC) study
- Primary biliary cirrhosis (PBC) is a chronic liver disease that leads to cirrhosis and eventually death

- 10-year study conducted by Mayo clinic (Murtagh et al., Hepatology, 1994) 

- 158 randomized to treatment, 154 to placebo

- Longitudinal biomarker measurements of serum bilirubin at times 0, 6m, 1y, 2y, etc.

\medskip
**Outcomes:**

  1. Longitudinal biomarker: serum bilirubin

  2. Time to death

**Question of interest:**

  - What is the association between the time-varying serum bilirubin (that is measured with error) and the risk of death?

```{r "pbc.data", eval=T, echo = T, include=T, out.width="90%"}
head(pbc2, 2)

```

## Joint model
```{r "jm", eval=T, echo = T, include=T, out.width="90%"}

lme.fit <- lme(log(serBilir) ~ year + year:drug, 
              data=pbc2, random = ~year|id)

surv.fit <- coxph(Surv(years, status2) ~ drug, 
              data=pbc2.id, x=TRUE)

## the knots of the piecewise constant are chosen based on the percentiles of events (5-6 internal knots by default)
joint.fit <- jointModel(lme.fit, surv.fit, 
              timeVar="year", method = "piecewise-PH-GH")
 
```

## Summary
```{r "jm.sum", eval=T, echo = T, include=T, out.width="90%"}
##summary(joint.fit)

plot(joint.fit)
```

## Confidence intervals
```{r "jm.ci", eval=T, echo = T, include=T, out.width="90%"}
confint(joint.fit, parm='all')

```

## Comparison of nested models
```{r "jm.lrt", eval=T, echo = T, include=T, out.width="90%"}
surv.fit2 <- coxph(Surv(years, status2) ~ 1, 
              data=pbc2.id, x=TRUE)

joint.fit2 <- jointModel(lme.fit, surv.fit2, 
              timeVar="year", method = "piecewise-PH-GH")

anova(joint.fit2, joint.fit)

```
There is no drug effect in the survival outcome (model fits about the same). 


# More practice with joint models
Use the data *aids* from the JM package. Take a look at *?aids* for a description of the dataset. 

A randomized clinical trial in which longitudinal and survival data was collected to compare the efficacy and safety of two antiretroviral drugs in treating patients who had failed/were intolerant of AZT therapy

- Longitudinal marker: CD4 cell counts
- Survival outcome: time to death
- Research goal: How is CD4 cell counts associated with a person’s risk of death?
- Lower CD4 cell counts are associated with increased risk of death

```{r "aids.data", eval=T, echo = T, include=T, out.width="90%"}
head(aids, 2)

```

### Fit the joint model based on the fitted linear mixed and Cox models using the function jointModel()

Interpret the association between CD4 counts and the risk of death.

