---
title: "BIOS6643. L05:  Full rank, Estimability and Contrasts"
cheauthor: "EJC"
-- date: "" 
header-includes:
   - \usepackage{amsmath}
   - \usepackage{float}
output: pdf_document
---

\newcommand{\bi}{\begin{itemize}}
\newcommand{\ei}{\end{itemize}}
\newcommand{\itt}{\item}


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- supplement of L05. Only R notes are included here -->


# Notes

```{r installp, echo = FALSE}
## install.packages("KMsurv")
## install.packages("km.ci")
## install.packages("survRM2")
library(dplyr)
library(tidyverse)
library(ggplot2)
library(nlme)

```


##
### Example 5: Prospective randomized trial

STEPPED-CARE randomized trial. The dataset we will use the class resembles the trial.

- A behavioral intervention was tested versus usual care in 286 patients with lung or head and neck cancer.  
- Population: low income patients in the Denver area across 5 hospitals
- Primary outcomes: anxiety, depression and coping skills scores
- Outcomes were measured at baseline, and at 6, 12 and 24 weeks


```{r "read stepped-care", eval=T, echo=TRUE, include=TRUE}
# Read in data 
dat.step <- read.csv("/Users/juarezce/Documents/OneDrive - The University of Colorado Denver/BIOS6643/BIOS6643_Notes/data/stepped-care-class.csv", header=TRUE)

head(dat.step, 3)
```



```{r test, eval=T, echo = F, fig.height=3.7, include=TRUE}
  ggplot(dat.step, aes(x = time, y = cops, group = id)) +
  # facet_wrap(~treat) +
  geom_line(aes(color = treat), alpha = 0.2) +
  theme_classic() +
  theme(legend.position = "top") +
  ylab("Copping Skill")

```

# Tests using the Stepped Care data

## 
1. Write a model that includes intervention arm and time as categorical/class variables, plus intervention $\times$ time interaction.

2. Provide estimates of means at 6 weeks for the control and intervention arms.

3. Provide a test for the mean difference at 6 weeks between the control and intervention arms.

4. Write a test to compare is if trends over time differ between intervention/treatment arms.

5. Write a test to compare changes from beginning to end among the 2 treatment arms.


## 1. Write a model that includes intervention arm and time as class variables, plus intervention $\times$ time interaction.

```{r models, eval=T, echo = T, include=TRUE}
dat.step$time <- as.factor(dat.step$time)

## parameterization 1
fit1 <- lme(cops ~  treat*time,
            random= ~ 1 | id, data = dat.step)
summary(fit1) 

## parameterization 1
fit2 <- lme(cops ~  time + treat:time-1,
            random= ~ 1 | id, data = dat.step)
summary(fit2) 

```
# Estimating means in the control and intervention arm at 6 weeks
```{r eval=T, echo = T, include=TRUE}
## estimate the means in the control and intervention arm at 6 weeks
coef2 <- summary(fit2)$coef$fixed
cov2 <- summary(fit2)$varFix

c6.c <- c(0, 1, 0, 0,
          0, 0, 0, 0)
c6.i <- c(0, 1, 0, 0,
          0, 1, 0, 0)
mu6.control <- c6.c %*% coef2 
mu6.int <- c6.i %*% coef2 

mu6.control
mu6.int

se6.control <- sqrt(c6.c %*% cov2  %*%  c6.c)
se6.int <- sqrt(c6.i %*% cov2  %*%  c6.i)
## mean, SE in control 
c(mu6.control, se6.control)
## mean, SE in intervention
c(mu6.int, se6.int)
```
# Estimating the mean difference in the control and intervention arm at 6 weeks
```{r eval=T, echo = T, include=TRUE}
## estimate the difference between intervention and control arm
c6.dif <- c(0, 0, 0, 0,
            0, 1, 0, 0)
mu6.dif <- c6.dif %*% coef2 
se6.dif <- sqrt(c6.dif %*% cov2  %*%  c6.dif)
c(mu6.dif, se6.dif)
```

# Test the hypothesis that there is an interaction between time and treatment arm
```{r eval=T, echo = T, include=TRUE}
## reduced model
fit.red <- lme(cops ~  time -1,
            random= ~ 1 | id, data = dat.step)
summary(fit.red) 
anova.lme(fit2, fit.red) 

fit2$logLik
fit.red$logLik
2*(fit2$logLik-fit.red$logLik)
pchisq(2*(fit2$logLik-fit.red$logLik), 4, ncp=0, lower.tail=FALSE, log.p=FALSE)
```
# Construct a wald-type test for the hypothesis that there is an interaction between time and treatment arm
```{r eval=T, echo = T, include=TRUE}
##  testing all parameters corresponding to the diff in means are 0
c.w <- matrix(c(0, 0, 0, 0, 1, 0, 0, 0,
                0, 0, 0, 0, 0, 1, 0, 0,
                0, 0, 0, 0, 0, 0, 1, 0,
                0, 0, 0, 0, 0, 0, 0, 1), byrow=T, 4, 8)

c.wa <- c.w%*%coef2
## chisquare test value
chisq.v <- t(c.wa)%*%solve(c.w%*% cov2  %*%t(c.w)) %*% c.wa 
pval.w <- pchisq(chisq.v, 4, ncp=0, lower.tail=FALSE, log.p=FALSE)
pval.w

```

