---
title: "BIOS6643. L10 Semiparametric regression"
cheauthor: "EJC"
-- date: "" 
header-includes:
   - \usepackage{amsmath}
   - \usepackage{float}
output: pdf_document
---


```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- -->


```{r installp, echo = FALSE, eval=T, include=F}
##library(dplyr)
library(tidyverse)
library(ggplot2)

library(lme4)
library(ISLR) ##
library(splines)

```

# Modeling non-linearity in crossectional data
## Example using the Wage data (James, Witten (2013 Ch 7)) - polynomial regression

```{r eval=T, echo = T, include=TRUE}
#  load data
head(Wage,3)

## fitting a polynomial regression of degree 4, using orthogonal polynomials 
fit <- lm(wage ~ poly(age,4), data=Wage)
coef(summary(fit))

## fitting a polynomial regression of degree 4, using powers of age directly
fit2 <- lm(wage ~ poly(age,4, raw=TRUE), data=Wage) ## raw=TRUE
coef(summary(fit2))

## using coefficients of age^X 
fit2a <- lm(wage ~ age + I(age^2) + I(age^3) + I(age^4), data=Wage) 
coef(summary(fit2a))

## predict the outcome at a grid of values age
agelims <- range(Wage$age)
agelims
age.grid <- seq(from=agelims[1],to=agelims[2])
preds <- predict(fit, newdata=list(age=age.grid), se=TRUE)
se.bands <- cbind(preds$fit+2*preds$se.fit, preds$fit-2*preds$se.fit)

par(mfrow=c(1,2),mar=c(4.5,4.5,1,1),oma=c(0,0,4,0)) 
plot(Wage$age, Wage$wage, xlim=agelims, cex=.5, col="darkgrey")
title("Degree-4 Polynomial",outer=T)
lines(age.grid, preds$fit, lwd=2,col="blue")
matlines(age.grid, se.bands, lwd=1,col="blue",lty=3)

## predictions are the same (no meaningful differences) regardless of which basis is used
preds2 <- predict(fit2, newdata=list(age=age.grid), se=TRUE)
max(abs(preds$fit-preds2$fit))
  
table(cut(Wage$age,4))
fit3 <- lm(wage ~cut(age,4),data=Wage)
coef(summary(fit3))
preds3 <- predict(fit3, newdata=list(age=age.grid), se=TRUE)
se.bands3 <- cbind(preds3$fit+2*preds3$se.fit, preds3$fit-2*preds3$se.fit)

par(mfrow=c(1,2),mar=c(4.5,4.5,1,1),oma=c(0,0,4,0)) 
plot(Wage$age, Wage$wage, xlim=agelims, cex=.5, col="darkgrey")
title("Degree-4 Polynomial",outer=T)
lines(age.grid, preds3$fit, lwd=2,col="blue")
matlines(age.grid, se.bands3, lwd=1,col="blue",lty=3)


```

## Wage data - regression splines
We need the library **splines**. The **bs()** function generates the entire matrix of basis functions for splines with the specified set of knots.

```{r eval=T, echo = T, include=TRUE}

bst <- bs(age.grid, knots=c(25,40,60),  ## cubic with degree: 3 (knots)+4=7 (including intercept)
	degree=3)

## plotting some basis
plot.ind <- TRUE
if (plot.ind==TRUE) { 
par(mfrow=c(1,1))
plot(age.grid, bst[,1],type='l')
lines(age.grid, bst[,2],col=2)
lines(age.grid, bst[,3],col=3)
lines(age.grid, bst[,4],col=4)
lines(age.grid, bst[,5],col=5)
lines(age.grid, bst[,6],col=6)
}

## number of basis
dim(bs(Wage$age,knots=c(25,40,60)))
## it's possible to specify the degrees of freedom, in which case knots are placed at uniform quantiles of data
dim(bs(Wage$age,df=6))
attr(bs(Wage$age,df=6),"knots")  
## bs also has a "degree" option which corresponds to the degree of the polynomial


fit4 <- lm(wage ~ bs(age, knots=c(25,40,60)),
          data=Wage)
pred4 <- predict(fit4,
                newdata=list(age=age.grid), se=T)
plot(Wage$age, Wage$wage,col="gray")
lines(age.grid, pred4$fit,lwd=2, col='blue')
lines(age.grid, pred4$fit+2*pred4$se,lty="dashed")
lines(age.grid, pred4$fit-2*pred4$se,lty="dashed")


fit5 <- lm(wage ~ ns(age, df=4),data=Wage)
pred5 <- predict(fit5, newdata=list(age=age.grid), se=T)
lines(age.grid, pred5$fit,col="red",lwd=2)
legend('topright', lty=c(1,1), col=c("blue", "red"), 
       c("Cubic Spline", "Natural Cubic Spline"))

```


## Wage data - smoothing splines

```{r eval=T, echo = T, include=TRUE}

plot(Wage$age, Wage$wage,xlim=agelims,
     cex=.5, col="darkgrey")
title("Smoothing Spline")
fit6 <- smooth.spline(Wage$age, Wage$wage, df=16)
fit6b <- smooth.spline(Wage$age, Wage$wage,cv=TRUE)
fit6$df
lines(fit6,col="red",lwd=2)
lines(fit6b,col="blue",lwd=2)
legend("topright",legend=c("16 DF","6.8 DF"),
col=c("red","blue"),lty=1,lwd=2,cex=.8)

```

## Wage data - local polynomial regression
We need the function **loess()**. 

```{r eval=T, echo = T, include=TRUE}

plot(Wage$age,Wage$wage,ylim=c(20,200),cex=.5,col="darkgrey")
title("Local Regression")
fit7 <- loess(wage ~age, span=.2, data=Wage)
fit7b <- loess(wage ~age, span=.5, data=Wage)
lines(age.grid,predict(fit7, data.frame(age=age.grid)), col="red",lwd=2)
lines(age.grid,predict(fit7b, data.frame(age=age.grid)), col="blue",lwd=2)
legend("topright",legend=c("Span=0.2","Span=0.5"), col=c("red","blue"),lty=1,lwd=2,cex=.8)

```

# Modeling non-linearity in longitudinal data

## Colorado CF FEV data
The data set gives characteristics of children patients with a diagnosis of Cystic Fibrosis (CF) who are patients at the Colorado Children’s Hospital. Data pull was conducted in Summer 2020. This dataset may only be used for BIOS6643. 

VARIABLE DESCRIPTIONS:  
id: patient ID	
race: race	
age: Age in years	
gender	: gender
fev: \% of predicted forced expiratory volume in 1 second

## Local polynomial regression

```{r eval=T, echo = T, include=TRUE}
#  Read in the data 

##  Read in the data 
dat <- read.csv("/Users/juarezce/Documents/OneDrive - The University of Colorado Denver/BIOS6643/BIOS6643_Notes/data/fev.csv",
                header=TRUE)
length(unique(dat$id)) ## number of individuals
head(dat,3)
summary(dat$age) ## age in years
summary(dat$fev)

## getting the age grid for fev
agelimsf <- range(dat$age)
agelimsf
age.gridf <- seq(from=agelimsf[1],to=agelimsf[2])

plot(dat$age, dat$fev, ylim=c(20,150), cex=.5, col="darkgrey")
title("Local Regression")
fitf.0a <- loess(fev ~age, span=.1, data=dat)
fitf.0b <- loess(fev ~age, span=.6, data=dat)
lines(age.gridf, predict(fitf.0a, data.frame(age=age.gridf), family="gaussian"),
      col="red",lwd=2)
lines(age.gridf, predict(fitf.0b, data.frame(age=age.gridf), family="gaussian"), 
      col="blue",lwd=2)
legend("topright",legend=c("Span=0.1","Span=0.6"), col=c("red","blue"),lty=1,lwd=2,cex=.8)

```

## Polynomial regression 

```{r eval=T, echo = T, include=TRUE}

fit.lmm.1 <- lmer(fev ~ poly(age, 7) + ( 1 | id), data=dat)
coef(summary(fit.lmm.1))

fit.lmm.1b <- lmer(fev ~ poly(age, 7, raw=TRUE) + ( 1 | id), data=dat)
coef(summary(fit.lmm.1b))

pred.1 <- predict(fit.lmm.1,
                  newdata= data.frame(age=age.gridf, poly(age.gridf, 7)),
                  type='response', re.form=~0)

par(mfrow=c(1,1),mar=c(4.5,4.5,1,1),oma=c(0,0,4,0)) 
plot(dat$age, dat$fev, xlim=agelimsf, cex=.5, col="darkgrey",
     xlab='Age', ylab='Fev1')
title("Degree-7 Polynomial",outer=T)
lines(age.gridf, pred.1, lwd=2,col="blue")

```

## Spline regression 

```{r eval=T, echo = T, include=TRUE}

fit.lmm.2 <- lmer(fev ~ bs(age, knots=c(10,15)) + ( 1 | id), data=dat)
coef(summary(fit.lmm.2))

fit.lmm.2b <- lmer(fev ~ ns(age, knots=c(10,15)) + ( 1 | id), data=dat)
coef(summary(fit.lmm.2b))

pred.2 <- predict(fit.lmm.2,
                  newdata= data.frame(age=age.gridf, bs(age.gridf, knots=c(10,15))),
                  type='response', re.form=~0)
pred.2b <- predict(fit.lmm.2,
                  newdata= data.frame(age=age.gridf, ns(age.gridf, knots=c(10,15))),
                  type='response', re.form=~0)

par(mfrow=c(1,1),mar=c(4.5,4.5,1,1),oma=c(0,0,4,0)) 
plot(dat$age, dat$fev, xlim=agelimsf, cex=.5, col="darkgrey",
     xlab='Age', ylab='Fev1')
title("Cubic spline",outer=T)
lines(age.gridf, pred.2,col="blue",lwd=2)
lines(age.gridf, pred.2b,col="red",lwd=2)
legend('topright', lty=c(1,1), col=c("blue", "red"), 
       c("Cubic Spline", "Natural Cubic Spline"))


```


## Piecewise linear mean form/structure  
Fitting piecewise linear model with a knot at 12 (or 14 years). The model is parameterized as 

$E(y_{ij})=\beta_0 + \beta_1 age_{ij} + \beta_2 (age_{ij} -12) I(age_{ij}>12)$

For age < 12: $E(y_{ij})=\beta_0 + \beta_1 age_{ij}$

For age $\geq$ 12: $E(y_{ij})=\beta_0 + \beta_1 age_{ij} + \beta_2 (age_{ij} -12) = (\beta_0 -12) + (\beta_1+\beta_2 ) age_{ij}$


```{r eval=T, echo = T, fig.height=3.7, include=T}

dat <- dat %>%
  mutate(age.12 = as.numeric(if_else(age > 12, age - 12 , 0)),
         age.14 = as.numeric(if_else(age > 14, age - 14 , 0)))
head(dat,3)

fit.lmm.3 <- lmer(fev ~ age + age.12 + ( 1 | id), data=dat)
summary(fit.lmm.3)

fit.lmm.3b <- lmer(fev ~ age + age.14 + ( 1 | id), data=dat)
summary(fit.lmm.3b)

pred.3 <- predict(fit.lmm.3, newdata=data.frame(age=age.gridf,
                                                age.12 = ifelse(age.gridf>12, age.gridf-12, 0)),
                  re.form=~0)
pred.3b <- predict(fit.lmm.3b, newdata=data.frame(age=age.gridf,
                                                  age.14 = ifelse(age.gridf>14, age.gridf-14, 0)),
                   re.form=~0)
par(mfrow=c(1,1),mar=c(4.5,4.5,1,1),oma=c(0,0,4,0)) 
plot(dat$age, dat$fev, xlim=agelimsf, cex=.5, col="darkgrey",
     xlab='Age', ylab='Fev1')
title("Piecewise linear",outer=T)
lines(age.gridf, pred.3,col="blue",lwd=2)
lines(age.gridf, pred.3b,col="red",lwd=2)
legend('topright', lty=c(1,1), col=c("blue", "red"), 
       c("PW Linear with Cutpoint at 12 years", "PW Linear with Cutpoint at 14 years"))


```


