---
title: "BIOS6643. L10 Model Building"
cheauthor: "EJC"
-- date: "" 
header-includes:
   - \usepackage{amsmath}
   - \usepackage{float}
output: pdf_document
---

\newcommand{\bi}{\begin{itemize}}
\newcommand{\ei}{\end{itemize}}
\newcommand{\itt}{\item}


```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- -->


```{r installp, echo = FALSE, eval=T, include=F}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(nlme)
library(lme4)

```

# Exercise involving $\pmb G$ and $\pmb R$ matrices

Consider a basic science experiment conducted where cell counts are measured at 4 time points for samples taken from individual subjects or animals.  A linear mixed model will be fit for the data (perhaps after log transformation), and fixed effects will be included for time, and treatment group as well as their interaction.  (To answer this question we do not need to know the specific form of $\pmb {X\beta}$.)  

Determine the structure for $\pmb V_i$ if a random intercept for subjects will be included, plus an AR(1) structure for the error covariance matrix ($\pmb R_i$).  What does the combination of non-simple $\pmb R$ and $\pmb G$ allow you to do in modeling covariance that using only one cannot do?  Discuss in a few sentences.

# FEV Study 

## Description
The data set gives characteristics of children patients with a diagnosis of Cystic Fibrosis (CF) who are patients at the Colorado Children’s Hospital. Data pull was conducted in Summer 2020. This dataset may only be used for BIOS6643. 

VARIABLE DESCRIPTIONS:  
id: patient ID	
race: race	
Genotypes1: mutation of copy one of the gene that codes for the CFTR protein
Genotypes2: mutation of copy two of the gene that codes for the CFTR protein. 
age: Age in years	
gender	: gender
fev: \% of predicted forced expiratory volume in 1 second

- **Objectives** of the study included:
  - Determine if fev values over time are larger on average for males than for females
  - Determine if the rate of change of fev over time is different for males and females.

## Selecting the mean structure

```{r eval=T, echo = T, include=TRUE}
#  Read in the data 

##  Read in the data 
dat <- read.csv("/Users/juarezce/Documents/OneDrive - The University of Colorado Denver/BIOS6643/BIOS6643_Notes/data/fev.csv",
                header=TRUE)
length(unique(dat$id)) ## number of individuals
head(dat,3)
summary(dat$age) ## age in years
summary(dat$fev)


######
################################################################################################
### selecting the mean form
## --------------------------------------------------------------------------------------------

ggplot(dat, aes(x = age, y = fev, group = id)) +
    facet_wrap(~gender) +
    geom_line(aes(color = gender), alpha = 0.3) +
    geom_smooth(aes(group = 1), method="loess", color="black", se=FALSE) +
    theme_classic() +
    theme(legend.position = "top") +
    ylab("Fev")

## age as linear
fit.lme.0 <- lme(fev ~ -1 + gender + age:gender, data=dat,
                    random = ~ 1 | id,method="ML")
summary(fit.lme.0)

## adding quadratic terms of age
fit.lme.1 <- lme(fev ~ -1 + gender  + age:gender + I(age*age):gender, 
                 data=dat,
                    random = ~ 1 | id,method="ML")
summary(fit.lme.1)

## adding cubic terms of age
fit.lme.2 <- lme(fev ~ -1 + gender  + age:gender + I(age*age):gender +
                   I(age*age*age):gender, data=dat,
                    random = ~ 1 | id,method="ML")
summary(fit.lme.2)

## updating model - same as above
## fit.lme.2 <- update(fit.lme.1, fev ~ -1 + gender  + age:gender + I(age*age):gender +
##                   I(age*age*age):gender)

anova(fit.lme.0, fit.lme.1, fit.lme.2)

```


## Preliminary covariance structure $\pmb G$

```{r eval=T, echo = T, fig.height=3.7, include=T}
################################################################################################
### 


################################################################################################
### preliminary covariance G
## --------------------------------------------------------------------------------------------

fit.lm.0 <- lm(fev ~ -1 + gender  + age:gender + I(age*age):gender + 
                 I(age*age*age):gender, data=dat)
summary(fit.lm.0)

## calculating residuals
dat$resid <- residuals(fit.lm.0)
head(dat)    

ggplot(dat, aes(x = age, y = resid, group = id)) +
    facet_wrap(~gender) +
    geom_line(aes(color = gender), alpha = 0.3) +
    geom_smooth(aes(group = 1), method="loess", color="black", se=FALSE) +
    theme_classic() +
    theme(legend.position = "top") +
    ylab("residuals")

ggplot(dat, aes(x = age, y = resid^2, group = id)) +
    facet_wrap(~gender) +
    geom_point(aes(color = gender), alpha = 0.3) +
    geom_smooth(aes(group = 1), method="loess", color="black", se=FALSE) +
    theme_classic() +
    theme(legend.position = "top") +
    ylab("Squared residuals")

## model with quadratic random effects did not converge: adding "+ I(age*age)" to random effects
fit.lme.3 <- lme(fev ~ -1 + gender  + age:gender + I(age*age):gender +
                   I(age*age*age):gender, data=dat,
                    random = ~ age  | id, method="ML")

summary(fit.lme.2)
summary(fit.lme.3)

```

## Residual covariance structure $\pmb R$

```{r eval=T, echo = T, fig.height=3.7, include=T}
################################################################################################
### 

## different variance for males and females
fit.lme.4 <- lme(fev ~ -1 + gender  + age:gender + I(age*age):gender +
                   I(age*age*age):gender, data=dat,
                    random = ~ age  | id, method="ML", 
                    weights = varIdent(form = ~ 1 | gender))
summary(fit.lme.4)
R.b.1 <- getVarCov(fit.lme.4,type="conditional",individual=1)   # R_1; first male
R.b.2 <- getVarCov(fit.lme.4,type="conditional",individual=12)   # R_2; first female
R.b.1
R.b.1

## allowing for AR(1) exp
fit.lme.5 <- lme(fev ~ -1 + gender  + age:gender + I(age*age):gender +
                   I(age*age*age):gender, data=dat,
                    random = ~ age  | id, method="ML", 
                   corExp(form = ~ age | id))
summary(fit.lme.5)

anova(fit.lme.3, fit.lme.4, fit.lme.5) 

```


## Model reduction 

```{r eval=T, echo = T, include=T}
## do we really random slopes? The variance of the random slopes, is not negligible (~1.8^2), so I would keep random slopes in


## testing if the cubic terms are needed in the mean form
fit.lme.5.1 <- lme(fev ~ -1 + gender  + age:gender + I(age*age):gender,
                 data=dat,
                 random = ~ age  | id, method="ML", 
                 corExp(form = ~ age | id))
summary(fit.lme.5.1)

anova(fit.lme.5, fit.lme.5.1) ## we do seem to need cubic terms

## testing if there is an interaction between age and gender


```

## Final model

```{r eval=T, echo = T, fig.height=5, include=T}



```


