---
title: "BIOS6643. L07:  Random effects"
cheauthor: "EJC"
-- date: "" 
header-includes:
   - \usepackage{amsmath}
   - \usepackage{float}
output: pdf_document
---

\newcommand{\bi}{\begin{itemize}}
\newcommand{\ei}{\end{itemize}}
\newcommand{\itt}{\item}


```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- supplement of L05. Only R notes are included here -->


# Notes

```{r installp, echo = FALSE, eval=T, include=T}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(nlme)
library(lme4)

```


##
### Example 5: Prospective randomized trial

STEPPED-CARE randomized trial. The dataset we will use the class resembles the trial.

- A behavioral intervention was tested versus usual care in 286 patients with lung or head and neck cancer.  
- Population: low income patients in the Denver area across 5 hospitals
- Primary outcomes: anxiety, depression and coping skills scores
- Outcomes were measured at baseline, and at 6, 12 and 24 weeks


```{r "read stepped-care", eval=T, echo=TRUE, include=TRUE}
# Read in data 
dat.step <- read.csv("/Users/juarezce/Documents/OneDrive - The University of Colorado Denver/BIOS6643/BIOS6643_Notes/data/stepped-care-class.csv", header=TRUE)

head(dat.step, 3)
```


```{r test, eval=T, echo = F, fig.height=3.7, include=TRUE}
  ggplot(dat.step, aes(x = time, y = cops, group = id)) +
  # facet_wrap(~treat) +
  geom_line(aes(color = treat), alpha = 0.2) +
  theme_classic() +
  theme(legend.position = "top") +
  ylab("Copping Skill")

```

# Using nlme package to fit models with the Stepped Care data

## 


```{r lme.models, eval=T, echo = T, include=TRUE}
## parameterization 1
dat.step$time <- as.factor(dat.step$time)

step.lme <- lme(cops ~  time + treat:time-1,
            random= ~ 1 | id, data = dat.step)
summary(step.lme) 

beta <- fixed.effects(step.lme)  #  may use fixef(step.lme)
b <- random.effects(step.lme)    #  random effects; may use ranef(step.lme)
head(b,3)
se.beta <- sqrt(diag(step.lme$varFix)) 
## Recall step.lme$varFix provides the var-cov of the fixed coefficients of model (beta)
cov <- step.lme$varFix

G <- getVarCov(step.lme, type="random.effects")   #  G matrix 
sigma2 <- step.lme$sigma^2  # sigma^2
R <- getVarCov(step.lme, type="conditional", individual=1)   # R_i
V <- getVarCov(step.lme, type="marginal", individual=1)   # V_i


```

# Using lme4 package to fit models with the Stepped Care data

## 
- **lme4** is more computationally efficient than **nlme** 
- **lme4** does not currently implement **nlme**’s features for modeling heteroscedasticity and cor-
relation of residuals.
- **lme4** includes generalized linear mixed model (GLMM) capabilities, via the **glmer** function.

```{r lmer.models, eval=T, echo = T, include=TRUE}
## parameterization 1
library(lme4) ## lme4 seems to be more computationally efficient than nlme
step.lmer <- lmer(cops ~ -1 + time + treat:time-1 + (1| id),
                      REML=FALSE,data=dat.step)
summary(step.lmer)

beta.lmer <- fixef(step.lmer) 
cov.lmer <- vcov(step.lmer) ## var-cov matrix of fixed coeff 

b.lmer <- ranef(step.lmer) ## random effects 
sigma2.lmer <- sigma(step.lmer)^2

vc <- VarCorr(step.lmer)

print(vc, comp="Variance")


```

```

