---
title: "BIOS6643. L14 Review exercises"
cheauthor: "EJC"
-- date: "" 
header-includes:
   - \usepackage{amsmath}
   - \usepackage{float}
output: pdf_document
---

\newcommand{\bi}{\begin{itemize}}
\newcommand{\ei}{\end{itemize}}
\newcommand{\itt}{\item}


```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- -->


```{r installp, echo = FALSE, eval=T, include=F}
##library(dplyr)
library(tidyverse)
library(multcomp)
library(nlme)
library(lme4)


```



# Question 1. Models for Beta Carotene Data

Using the Beta carotene dataset, answer the following questions using a mixed model with no random effects but an unstructure covariance R.  

  a. Estimate the mean response (beta carotene level) at all weeks in Roche (group2) and BASF 30mg (group 3). 

  b. Test for an interaction between Roche (group2) and BASF 30mg (group 3).
  
  c. Test the hypothesis that the 4 groups differ at 6 weeks 


```{r "beta_mod1"}
## import data beta
beta <- read.csv("/Users/juarezce/Documents/OneDrive - The University of Colorado Denver/BIOS6643/BIOS6643_Notes/data/beta_carotene_univar.csv",
               header=TRUE) 
head(beta,3)
names(beta) <- c("prepar", "id", "y", "time")
#  mutate(time = as.integer(time))

mod1 <- gls(y ~ 1 + factor(time) * factor(prepar), 
            ## UN for correlation! not covariance
            correlation = corSymm(form = ~1|id),
            ## for unequal variance over time
            weights = varIdent(form = ~1|time),
            method = "REML",
            data = beta)
```

\normalsize

\textcolor{blue}{a. Estimate the mean response (beta carotene level) at all weeks in group 2 and in group 3. }

```{r "get_contrast"}
## fixed effects
# beta_hat <- fixed.effects(mod1)
beta_hat <- coef(mod1)
## fixed effects variance-covariance
covb <- vcov(mod1) 

## mean group 2
mu2 <- matrix(NA, 5, 2)
## mean  group 3
mu3 <- matrix(NA, 5, 2)

##             int   time           prepar    prepar=2*ti    prepar=3*ti     prepar=4*ti
c21  <-     c( 1,   0, 0, 0, 0,   1, 0, 0,    0, 0, 0, 0,    0, 0, 0, 0,     0, 0, 0, 0) 
c22  <-     c( 1,   1, 0, 0, 0,   1, 0, 0,    1, 0, 0, 0,    0, 0, 0, 0,     0, 0, 0, 0) 
c23  <-     c( 1,   0, 1, 0, 0,   1, 0, 0,    0, 1, 0, 0,    0, 0, 0, 0,     0, 0, 0, 0) 
c24  <-     c( 1,   0, 0, 1, 0,   1, 0, 0,    0, 0, 1, 0,    0, 0, 0, 0,     0, 0, 0, 0) 
c25  <-     c( 1,   0, 0, 0, 1,   1, 0, 0,    0, 0, 0, 1,    0, 0, 0, 0,     0, 0, 0, 0) 

##             int   time           prepar    prepar=2*ti    prepar=3*ti     prepar=4*ti
c31  <-     c( 1,   0, 0, 0, 0,   0, 1, 0,    0, 0, 0, 0,    0, 0, 0, 0,     0, 0, 0, 0) 
c32  <-     c( 1,   1, 0, 0, 0,   0, 1, 0,    0, 0, 0, 0,    1, 0, 0, 0,     0, 0, 0, 0) 
c33  <-     c( 1,   0, 1, 0, 0,   0, 1, 0,    0, 0, 0, 0,    0, 1, 0, 0,     0, 0, 0, 0) 
c34  <-     c( 1,   0, 0, 1, 0,   0, 1, 0,    0, 0, 0, 0,    0, 0, 1, 0,     0, 0, 0, 0) 
c35  <-     c( 1,   0, 0, 0, 1,   0, 1, 0,    0, 0, 0, 0,    0, 0, 0, 1,     0, 0, 0, 0) 

## calculating estimates of linear combination
## function to calculate linear combination  of beta parameters: est and SE (standard error)
## input: lvec, vector to calculate linear combination
## output: c(est, SE), point estimate and SE
f.lbeta <- function(lvec){
 est <- t(lvec) %*% beta_hat  
 se <- sqrt(t(lvec) %*% covb %*% lvec)
 return(c(est, se))
}

mu2[1,] <- f.lbeta(c21)
mu2[2,] <- f.lbeta(c22)
mu2[3,] <- f.lbeta(c23)
mu2[4,] <- f.lbeta(c24)
mu2[5,] <- f.lbeta(c25)
## MEANs for group 2
mu2

mu3[1,] <- f.lbeta(c31)
mu3[2,] <- f.lbeta(c32)
mu3[3,] <- f.lbeta(c33)
mu3[4,] <- f.lbeta(c34)
mu3[5,] <- f.lbeta(c35)
## MEANs for group 3
mu3

```
The means for group are higher. Note we could plot them and typically accompany the interaction test in b with a figure or table of these means (or the mean difference).   

\normalsize

\newpage
\textcolor{blue}{  b. Test for an interaction between Roche (group2) and BASF 30mg.}
To answer this question we may think about the differences between each time point versus baseline and then the differences between groups. 

```{r "contrast"}



contr0 <- cbind(c22-c21-(c32-c31), c23-c21-(c33-c31), c24-c21-(c34-c31), c25-c21-(c35-c31))
t(contr0)

## contrast point estimates to be
ce0 <- t(contr0) %*% beta_hat
## contrast variance covariance matrix
cov0 <- t(contr0) %*% covb %*% contr0

## 
W0 <- t(ce0) %*% solve(cov0) %*% ce0
pchisq(W0, df = 4, lower.tail = FALSE)


## emmeans is a package cover 
test1 <- multcomp::glht(mod1, t(contr0))
summary(test1, test = Chisqtest())
```

\normalsize
There is a significant interaction between Roche (group2) and BASF 30mg. 






c.	Conduct a test to compare if the 6 week - baseline value differs between the 4 $groups$.



```{r}

##             int   time           prepar    prepar=2*ti    prepar=3*ti     prepar=4*ti
c11  <-     c( 1,   0, 0, 0, 0,   0, 0, 0,    0, 0, 0, 0,    0, 0, 0, 0,     0, 0, 0, 0) 
c12  <-     c( 1,   1, 0, 0, 0,   0, 0, 0,    0, 0, 0, 0,    0, 0, 0, 0,     0, 0, 0, 0) 
c13  <-     c( 1,   0, 1, 0, 0,   0, 0, 0,    0, 0, 0, 0,    0, 0, 0, 0,     0, 0, 0, 0) 
c14  <-     c( 1,   0, 0, 1, 0,   0, 0, 0,    0, 0, 0, 0,    0, 0, 0, 0,     0, 0, 0, 0) 
c15  <-     c( 1,   0, 0, 0, 1,   0, 0, 0,    0, 0, 0, 0,    0, 0, 0, 0,     0, 0, 0, 0) 

##             int   time           prepar    prepar=2*ti    prepar=3*ti     prepar=4*ti
c41  <-     c( 1,   0, 0, 0, 0,   0, 0, 1,    0, 0, 0, 0,    0, 0, 0, 0,     0, 0, 0, 0) 
c42  <-     c( 1,   1, 0, 0, 0,   0, 0, 1,    0, 0, 0, 0,    0, 0, 0, 0,     1, 0, 0, 0) 
c43  <-     c( 1,   0, 1, 0, 0,   0, 0, 1,    0, 0, 0, 0,    0, 0, 0, 0,     0, 1, 0, 0) 
c44  <-     c( 1,   0, 0, 1, 0,   0, 0, 1,    0, 0, 0, 0,    0, 0, 0, 0,     0, 0, 1, 0) 
c45  <-     c( 1,   0, 0, 0, 1,   0, 0, 1,    0, 0, 0, 0,    0, 0, 0, 0,     0, 0, 0, 1) 

## mean group 2
mu1 <- matrix(NA, 5, 2)
## mean  group 3
mu4 <- matrix(NA, 5, 2)

mu1[1,] <- f.lbeta(c11)
mu1[2,] <- f.lbeta(c12)
mu1[3,] <- f.lbeta(c13)
mu1[4,] <- f.lbeta(c14)
mu1[5,] <- f.lbeta(c15)
## MEANs for group 1
mu1

mu4[1,] <- f.lbeta(c41)
mu4[2,] <- f.lbeta(c42)
mu4[3,] <- f.lbeta(c43)
mu4[4,] <- f.lbeta(c44)
mu4[5,] <- f.lbeta(c45)
## MEANs for group 4
mu4

contr2 <- cbind(c25-c21 - (c15 -c11), 
                c35-c31 - (c15 -c11), 
                c45-c41 - (c15 -c11))
## contrast point estimates to be
(ce2 <- t(contr2) %*% beta_hat)
## contrast variance covariance matrix
cov2 <- t(contr2) %*% covb %*% contr2

## with both point estimates and standard deviation 
## an anova or pairwise comparison can be performed
W2 <- t(ce2) %*% solve(cov2) %*% ce2
pchisq(W2, df = 3, lower.tail = FALSE)
```


```{r "contrast2_glht"}
## emmeans is a package cover 
test2 <- multcomp::glht(mod1, t(contr2))
summary(test2, test = Chisqtest())
```


# Question 2. Models for Stepped Care Data
## STEPPED-CARE randomized trial
The dataset we will use in class resembles the trial.

- A behavioral intervention was tested versus usual care in 286 patients with lung or head and neck cancer.  
- Population: low income patients in the Denver area across 5 hospitals
- Primary outcomes: anxiety, depression and coping skills scores
- Outcomes were measured at baseline, and at 6, 12 and 24 weeks

Consider the stepped care data. Use a linear mixed model with time as continuous variable and random intercepts and random slopes. Assume there are no differences in coping self-efficacy score (CSES) at baseline. 

  a. Estimate the mean CSES at 6 weeks for both intervention groups. 
  
  b. Estimate the mean CSES difference at 6 weeks. 
  
  c. Test the hypothesis that the mean difference 12 weeks - baseline differs across the two treatment groups.
  



```{r "read stepped-care", eval=T, echo=TRUE, include=TRUE}
# Read in data 
dat.step <- read.csv("/Users/juarezce/Documents/OneDrive - The University of Colorado Denver/BIOS6643/BIOS6643_Notes/data/stepped-care-class.csv", header=TRUE)

head(dat.step, 3)



```

