---
title: "BIOS6643. L9 Specifying LMM through $G_i$ and $R_i$ structures"
cheauthor: "EJC"
-- date: "" 
header-includes:
   - \usepackage{amsmath}
   - \usepackage{float}
output: pdf_document
---

\newcommand{\bi}{\begin{itemize}}
\newcommand{\ei}{\end{itemize}}
\newcommand{\itt}{\item}


```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- -->


```{r installp, echo = FALSE, eval=T, include=F}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(nlme)
library(lme4)

```

# Dental study

## Description

- The orthodontic study data of Potthoff and Roy (1964). 

- World famous data set that is used to introduce features of longitudinal data modeling and analysis

- A study was conducted involving 27 children, 16 boys and 11 girls

- For each child, the distance (mm) from the center of the pituitary to the pterygomaxillary fissure was measured at ages 8, 10, 12, and 14 years of age 

- The pterygomaxillary fissure is a vertical opening in the human skull.

- **Objectives** of the study included:

  - Determine if distances over time are larger on average for boys than for girls
  - Determine if the rate of change of distance over time is different for boys and girls.


```{r eval=T, echo = T, include=TRUE}
#  Read in the data 
dat.den <- read.table("/Users/juarezce/Documents/OneDrive - The University of Colorado Denver/BIOS6643/BIOS6643_Notes/data/dental.txt")
dat.den <- dat.den[,2:5]      #  remove the first column
colnames(dat.den) <- c("id","age","distance","gender")

#  Total number of individuals
m <- max(dat.den$id)

head(dat.den, 2)
table(dat.den$gender)
```


## Some descriptives:  

```{r eval=T, echo = F, fig.height=3.7, include=T}
################################################################################################
### 

dat.den <- within(dat.den, {  id <- factor(id)
    gender <- factor(gender,levels=0:1,labels=c("Girls","Boys"))
})

ggplot(dat.den, aes(x = age, y = distance, group = id)) +
  # facet_wrap(~treat) +
  geom_line(aes(color = gender), alpha = 0.5) +
  theme_classic() +
  theme(legend.position = "top") +
  ylab("Distance")

```



## a. Common G matrix for both genders with random intercept and slopes    

```{r eval=T, echo = T, include=T}
#  (a) Common G matrix for both genders, default diagonal within-child                               
#      covariance matrix R_i with same variance sigma^2 for each
#      gender.  

dental.lme.a <- lme(distance ~ -1 + gender + age:gender,data=dat.den,
                    random = ~ age | id,method="ML")
summary(dental.lme.a)
beta.a <- fixed.effects(dental.lme.a)  #  beta, also fixef(dental.lme.a)
b.a <- random.effects(dental.lme.a)    #  posterior modes bi, also ranef(dental.lme.a)
sebeta.a <- summary(dental.lme.a)$tTable[,"Std.Error"]  
## Recall we can get the var-cov of fixed coefficients using 'varFix'
## dental.lme.a$varFix

G.a <- getVarCov(dental.lme.a, type="random.effects")   #  G matrix
sigma2.a <- dental.lme.a$sigma^2  # sigma^2
V.a <- getVarCov(dental.lme.a, type="marginal", individual=1)   # V_i
R.a <- getVarCov(dental.lme.a, type="conditional", individual=1)   # R_i
G.a
V.a
R.a

```

## b. Common G matrix with random intercepts and slopes, diagonal within-child covariance matrix R_i with different variance for each gender         
```{r eval=T, echo = T, include=T}

dental.lme.b <- lme(distance ~ -1 + gender + age:gender,data=dat.den,
                    random = ~ age | id, weights = varIdent(form = ~ 1 | gender),
                    method="ML")
beta.b <- fixed.effects(dental.lme.b)  #  beta
sebeta.model.b <-  sqrt(diag(dental.lme.b$varFix))
b.b <- random.effects(dental.lme.b)    #  posterior modes bi
G.b <- getVarCov(dental.lme.b, type="random.effects")   #  G matrix
R.b.1 <- getVarCov(dental.lme.b,type="conditional",individual=1)   # R_1; first girl
R.b.12 <- getVarCov(dental.lme.b,type="conditional",individual=12)   # R_12; first boy
G.b
R.b.1
R.b.12

#  As with gls(), when a weight statement is used, the standard
#  deviation for the first group is sigma, and those for other groups
#  are parameterized as this standard deviation x a factor. To get 
#  sigma^2_B, we must extract that factor from the varStruct object:
#
dental.lme.b$modelStruct$varStruct

sigma2vec.b <- matrix((1/unique(attributes(dental.lme.b$modelStruct$varStruct)$weights)*dental.lme.b$sigma)^2,
                      nrow=1,byrow=TRUE)
colnames(sigma2vec.b) <- c("sigma2.b.G","sigma2.b.B")
sigma2vec.b

V.b.1 <- getVarCov(dental.lme.b,type="marginal",individual=1)   # V_1 (girl)
V.b.12 <- getVarCov(dental.lme.b,type="marginal",individual=12)   # V_12 (boy)
V.b.1
V.b.12

```

## c. Common G matrix for both genders, and common within-child AR(1)

```{r eval=T, echo = T, include=T}

## Note that we cannot fit a model with separate G matrices for each gender

dental.lme.c <- lme(distance ~ -1 + gender + age:gender, data=dat.den,
                    random = ~ age | id,
                    correlation=corAR1(form = ~ age | id),
                    method="ML")
summary(dental.lme.c)
beta.c <- fixed.effects(dental.lme.c)
b.c <- random.effects(dental.lme.c)
sebeta.d <- sqrt(diag(dental.lme.c$varFix))
G.c <- getVarCov(dental.lme.c, type="random.effects")   #  G
sigma2.c <- dental.lme.c$sigma^2  # sigma^2
V.c <- getVarCov(dental.lme.c,type="marginal", individual=1)   # V_i
R.c <- getVarCov(dental.lme.c,type="conditional", individual=1)   # R_i
G.c
V.c
R.c

##  The correlation parameter alpha, which is constrained to be |alpha|<=1,
##  is estimated to be 0 (= Phi1).  So this fit is the same as (a)


```


## d. Common G matrix for both genders, common within-child AR(1) (exponential correlation) 
```{r eval=T, echo = T, include=T}

dental.lme.d <- lme(distance ~ -1 + gender + age:gender, data=dat.den,
                    random = ~ age | id,
                    correlation=corExp(form = ~ age | id),
                    method="ML")
summary(dental.lme.d)
beta.d <- fixed.effects(dental.lme.d)
b.d <- random.effects(dental.lme.d)
sebeta.model.d <-  sqrt(diag(dental.lme.d$varFix))
G.d <- getVarCov(dental.lme.d, type="random.effects")   #  G
sigma2.d <- dental.lme.d$sigma^2  # sigma^2
R.d <- getVarCov(dental.lme.d,type="conditional",individual=1)  #  R_i
V.d <- getVarCov(dental.lme.d,type="marginal",individual=1)   # V_i
R.d
V.d

```

## Comparison of models

```{r eval=T, echo = T, include=T}


## Compare the fitted models via AIC and BIC; model (b), with
#  diagonal R_i with gender-specific within-child variances and common
#  random intercept and slopes covariance matrix G is preferred

anova(dental.lme.a, dental.lme.b, dental.lme.c, dental.lme.d)

```


## Refitting final model

```{r eval=T, echo = T, include=T, fig.height=4}
##  Refit model (b) using REML and get 
dental.lme.b.reml <- lme(distance ~ -1 + gender + age:gender, data=dat.den,
                    random = ~ age | id, weights = varIdent(form = ~ 1 | gender))
beta.b.reml <- fixed.effects(dental.lme.b.reml)  #  beta
sebeta.model.b.reml <-  sqrt(diag(dental.lme.b$varFix))
b.b.reml <- random.effects(dental.lme.b.reml)    #  posterior modes bi
G.b.reml <- getVarCov(dental.lme.b.reml, type="random.effects")   #  G
G.b.reml

##  Random effects empirical Bayes estimates bhat_i for first 5 girls
b.b.reml[1:5,]

## PA predicted values $X_i \times betahat$ are produced by level=0;
## SS predicted values X_i betahat + Z_i bhat_i are produced by level =  1;
## both are gotten by level = 0:1

fitted(dental.lme.b.reml,level=0:1)[1:5,] 

## Can also extract both marginal (PA) residuals  Y_i-X_i betahat,
## and conditional (SS) residuals Y_i-X_i betahat - Z_i bhat_i
## level = 0 gives PA, level = 1 gives SS (default), and level = 0:1 gives both;
## type = default is "raw" residuals; can get standardized "pearson" (type=pearson)

residuals(dental.lme.b.reml, level=0:1)[1:5, ] 

###   PA and SS pearson residuals -- the PA residuals differ from those
##   obtained from PROC MIXED apparently due to different standardization
residuals(dental.lme.b.reml, level=0:1, type="pearson")[1:5, ] 


##  Plot SS residuals vs. predicted values

plot(dental.lme.b.reml, resid(. , type="p",level=1) ~ fitted(.,level=1) )

#  QQ plot of SS residuals 
qqnorm(dental.lme.b.reml, ~ resid(. , type="p",level=1),abline=c(0,1))

#  One can also make QQ plots and histograms of the bhat_i themselves
#  to assess the normality of the random effects, but remember that 
#  these are "shrunken" so could be misleading.  

qqnorm(dental.lme.b.reml, ~ ranef(.))

##  histograms of random effects
par(mfrow=c(1,2))
hist(b.b.reml[,1],xlab="Intercept Random Effect",main="Empirical Bayes Intercepts",freq=FALSE)
hist(b.b.reml[,2],xlab="Slope Random Effect",main="Empirical Bayes Slopes",freq=FALSE)

```

\newpage
# Fitting models using lme4:lmer

```{r eval=T, echo = T, include=T}

#  We fit only model (a), Common G matrix for both genders, default
#  diagonal wintin-child covariance matrix  R_i witih same variance
#  sigma^2 for each gender.  We use ML as in the lme() and SAS programs.

#  The random effects structure is specified in parentheses -- here,
#  we allow for random intercept and slope that are correlated

dental.lmer.a <- lmer(distance ~ -1 + gender + age:gender + (1 + age | id),
                      REML=FALSE, data=dat.den)

summary(dental.lmer.a)
## sebeta.model.a
##     gender0     gender1 gender0:age gender1:age 
##  1.18202362  0.98008221  0.09980390  0.08275303 

beta.lmer.a <- fixef(dental.lmer.a)
beta.lmer.a
b.lmer.a <- ranef(dental.lmer.a)
sigma2.lmer.a <- sigma(dental.lmer.a)^2

#  It is pretty unwieldy to extract the covariance matrix D of the
#  random effects.  We can look at the variances and correlation with
vc.a <- VarCorr(dental.lmer.a)

print(vc.a,comp="Variance")

#  All the covariance matrix stuff can be put in a data frame, from
#  which it can be extracted to form the matrices G, R_i, and V_i
vc.da <- as.data.frame(vc.a, order="lower.tri")
vc.da

G.lmer.a <- matrix(c(vc.da[1,4],vc.da[2,4],vc.da[2,4],vc.da[3,4]),2,2,
                   byrow=TRUE)
G.lmer.a

```

