---
title: "BIOS6643. L8 Covariance Structures $R_i$"
cheauthor: "EJC"
-- date: "" 
header-includes:
   - \usepackage{amsmath}
   - \usepackage{float}
output: pdf_document
---

\newcommand{\bi}{\begin{itemize}}
\newcommand{\ei}{\end{itemize}}
\newcommand{\itt}{\item}


```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- -->


```{r installp, echo = FALSE, eval=T, include=F}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(nlme)

```

# 1. Dental study

##
### Dental study

- The orthodontic study data of Potthoff and Roy (1964). 

- World famous data set that is used to introduce features of longitudinal data modeling and analysis

- A study was conducted involving 27 children, 16 boys and 11 girls

- For each child, the distance (mm) from the center of the pituitary to the pterygomaxillary fissure was measured at ages 8, 10, 12, and 14 years of age 

- The pterygomaxillary fissure is a vertical opening in the human skull.

- **Objectives** of the study included:

  - Determine if distances over time are larger on average for boys than for girls
  - Determine if the rate of change of distance over time is different for boys and girls.


```{r eval=T, echo = T, include=TRUE}
#  Read in the data 
dat.den <- read.table("/Users/juarezce/Documents/OneDrive - The University of Colorado Denver/BIOS6643/BIOS6643_Notes/data/dental.txt")
dat.den <- dat.den[,2:5]      #  remove the first column
colnames(dat.den) <- c("id","age","distance","gender")

#  Total number of individuals
m <- max(dat.den$id)

head(dat.den, 2)
table(dat.den$gender)
```

# Some descriptives statistics:  

```{r eval=T, echo = F, fig.height=3.7, include=T}
################################################################################################
### 

dat.den <- within(dat.den, {  id <- factor(id)
    gender <- factor(gender,levels=0:1,labels=c("Girls","Boys"))
})

ggplot(dat.den, aes(x = age, y = distance, group = id)) +
  # facet_wrap(~treat) +
  geom_line(aes(color = gender), alpha = 0.5) +
  theme_classic() +
  theme(legend.position = "top") +
  ylab("Distance")

################################################################################################
### 
## --------------------------------------------------------------------------------------------
##  Calculating covariance and correlation matrices

#  Calculate girl- and boy-specific overall sample covariance and correlation
#  matrices and inspect to get a sense of overall pattern of variation and
#  correlation

#  Reconfigure the data set from one record per observation to one record
#  per child using the reshape function (so from "long" to "wide" format 
#  in R-speak) 

dat.denw <- reshape(dat.den, v.names="distance", idvar="id",
                       timevar="age", direction="wide")
head(dat.denw,3)

##  Extract boys and girls separately
girlsdat <- dat.denw[dat.denw$gender=='Girls',]
boysdat <- dat.denw[dat.denw$gender=='Boys',]

##  Numbers of girls and boys
m.girls <- nrow(girlsdat)
m.boys <- nrow(boysdat)

#  Sample mean vectors -- we could instead fit a straight line to
#  the data on all m = 27 children and calculate the means as the predicted
#  values at 8, 10, 12, and 14 years
girls.mean <- apply(girlsdat[,3:6],2,mean)
boys.mean <- apply(boysdat[,3:6],2,mean)
girls.mean
boys.mean

#  Calculate the sample covariance matrix and correlation matrix for
#  each gender

girls.cov <- cov(girlsdat[,3:6])
boys.cov <- cov(boysdat[,3:6])

girls.corr <- cor(girlsdat[,3:6])  #  or use cov2cor(girls.cov)
boys.corr <- cor(boysdat[,3:6])

girls.cov
boys.cov

girls.corr
boys.corr
```
# Calculating autocorrelation:  

```{r eval=T, echo = F, fig.height=6, include=T}
################################################################################################
### 

##  Calculating covariance and correlation matrices

##  Center and scale to create scatterplot matrices using the scale and
##  pairs functions

girls.sd <- sqrt(diag(girls.cov))
boys.sd <- sqrt(diag(boys.cov))

girls.centscale <- scale(girlsdat[,3:6], center=girls.mean, scale=girls.sd)
boys.centscale <- scale(boysdat[,3:6], center=boys.mean, scale=boys.sd)

##  Now create the scatterplots

agelabs <- c("Age 8", "Age 10", "Age 12", "Age 14")

## scatterplot for girls
pairs(girls.centscale,label=agelabs,oma=c(5,5,5,5),main="Girls")

## scatterplot for boys
##pairs(boys.centscale,label=agelabs,oma=c(5,5,5,5),main="Boys")

#  Assuming a common pattern of overall covariance/correlation for 
#  boys and girls (but different overall means), calculate the pooled
#  covariance matrix and associated correlation matrix

cov.pooled <- ( (m.girls-1)*girls.cov + (m.boys-1)*boys.cov )/(m-2)
corr.pooled <- cov2cor(cov.pooled)
cov.pooled
corr.pooled

##  Autocorrelations at lags 1, 2, and 3 under stationarity for girls
ac.lag1 <- cor(c(girls.centscale[,1],girls.centscale[,2],girls.centscale[,3]),
           c(girls.centscale[,2],girls.centscale[,3],girls.centscale[,4]))
ac.lag2 <- cor(c(girls.centscale[,1],girls.centscale[,2]),
           c(girls.centscale[,3],girls.centscale[,4]))
ac.lag3 <- cor(girls.centscale[,1],girls.centscale[,4])

## autocorrelations
round(c(ac.lag1,ac.lag2,ac.lag3),3)

##  Create lag plots for girls
par(mfrow=c(3,1))
plot(c(girls.centscale[,1],girls.centscale[,2],girls.centscale[,3]),
     c(girls.centscale[,2],girls.centscale[,3],girls.centscale[,4]),
     xlab="at age",ylab="at age + 2 years",xlim=c(-2,2),ylim=c(-2,2))
title("Lag 1 (2 years)")

plot(c(girls.centscale[,1],girls.centscale[,2]),
     c(girls.centscale[,3],girls.centscale[,4]),xlab="at age",
     ylab="at age + 4 years",xlim=c(-2,2),ylim=c(-2,2))
title("Lag 2 (4 years)")

plot(girls.centscale[,1],girls.centscale[,4],xlab="at age",
     ylab="at age + 6 years",xlim=c(-2,2),ylim=c(-2,2))
title("Lag 3 (6 years)")

#  Autocorrelations at lags 1, 2, and 3 under stationarity for boys
#  (could be plotted as above)

ac.lag1.boys <- cor(c(boys.centscale[,1],boys.centscale[,2],boys.centscale[,3]),
           c(boys.centscale[,2],boys.centscale[,3],boys.centscale[,4]))
ac.lag2.boys <- cor(c(boys.centscale[,1],boys.centscale[,2]),
           c(boys.centscale[,3],boys.centscale[,4]))
ac.lag3.boys <- cor(boys.centscale[,1],boys.centscale[,4])

round(c(ac.lag1.boys,ac.lag2.boys,ac.lag3.boys),3)


```


## 1. Unstructured correlation

```{r eval=T, echo = T, include=T}

#  Common unstructured correlation with variances changing over time
#  for both genders -- we extract components of the fit.  

head(dat.den, 2)
dental.un <- gls(distance ~ -1 + gender + age:gender, data=dat.den,
                 correlation=corSymm(form =  ~ 1 | id),
                 weights = varIdent(form = ~ 1 | age), method="ML")
beta.un <- coef(dental.un)
sebeta.un <- summary(dental.un)$tTable[,"Std.Error"]

V.un <- getVarCov(dental.un)  ## V_i matrix
Gamma.un <- cov2cor(V.un) ## Gamma_i=correlation matrix
vars.un <- diag(V.un) ## Lambda_i= diagonal matrix of variances
V.un
Gamma.un
vars.un

```
## 2. Common Compound symmetry
```{r eval=T, echo = T, include=T}

dental.cs <- gls(distance ~ -1 + gender + age:gender, data=dat.den,
                 correlation=corCompSymm(form = ~ 1 | id),method="ML")
summary(dental.cs)
beta.cs <- coef(dental.cs)
sebeta.cs <- summary(dental.cs)$tTable[,"Std.Error"]
V.cs <- getVarCov(dental.cs)  
Gamma.cs <- cov2cor(V.cs)     
vars.cs <- diag(V.cs)
V.cs
Gamma.cs
vars.cs

```

## 3. Compound symmetry by gender

```{r eval=T, echo = T, include=TRUE}

dental.cs2 <- gls(distance ~ -1 + gender + gender:age, data=dat.den,
                  correlation=corCompSymm(form = ~ 1 | id),
                 weights = varIdent(form = ~ 1 | gender), method="ML")
beta.cs2 <- coef(dental.cs2)
sebeta.cs2 <- summary(dental.cs2)$tTable[,"Std.Error"]
V.cs2.girl <-  getVarCov(dental.cs2, individual=1)
V.cs2.boy <-  getVarCov(dental.cs2, individual=12)
Gamma.cs2.girl <- cov2cor(V.cs2.girl)
Gamma.cs2.boy <- cov2cor(V.cs2.boy)
V.cs2.girl
Gamma.cs2.girl
V.cs2.boy
Gamma.cs2.boy


```

## 4. AR(1)

```{r eval=T, echo = T, include=TRUE}

dental.ar1 <- gls(distance ~ -1 + gender + age:gender, data=dat.den,
                  correlation=corAR1(form = ~ 1 | id), method="ML")
beta.ar1 <- coef(dental.ar1)
sebeta.ar1 <- summary(dental.ar1)$tTable[,"Std.Error"]
V.ar1 <- getVarCov(dental.ar1)  #  or corMatrix(dental.un$modelStruct$corStruct)[[1]]
Gamma.ar1 <- cov2cor(V.ar1)
vars.ar1 <- diag(V.ar1)
V.ar1
Gamma.ar1
vars.ar1


```

## 5. One-dependent (banded)

```{r eval=T, echo = T, include=TRUE}

## autoregressive moving average process, with arbitrary orders for the autoregressive and moving average components.

dental.1d <- gls(distance ~ -1 + gender +  age:gender, data=dat.den,
                 correlation=corARMA(form = ~ 1 | id, q=1),method="ML")
beta.1d <- coef(dental.1d)
sebeta.1d <- summary(dental.1d)$tTable[,"Std.Error"]
V.1d <- getVarCov(dental.1d)  #  or corMatrix(dental.un$modelStruct$corStruct)[[1]]
Gamma.1d <- cov2cor(V.1d)
vars.1d <- diag(V.1d)
V.1d
Gamma.1d
vars.1d


```

## 6. Exponential spatial correlation

```{r eval=T, echo = T, include=TRUE}

## exponential spatial correlation
dental.exp <- gls(distance ~ -1 + gender +  age:gender, data=dat.den,
                 correlation=corExp(form = ~ 1 | id),method="ML")
V.exp <- getVarCov(dental.exp)  #  or corMatrix(dental.un$modelStruct$corStruct)[[1]]
Gamma.exp <- cov2cor(V.exp)
vars.exp <- diag(V.exp)
V.exp
Gamma.exp
vars.exp


```

## Comparing models

```{r eval=T, echo = T, include=TRUE}

anova(dental.un,dental.cs,dental.cs2,dental.ar1, dental.1d, dental.exp)


#  We now fit full and reduced models with this covariance structure
#  and compare via likelihood ratio test

full <- gls(distance ~ -1 + gender + gender:age, data = dat.den,
            correlation=corCompSymm(form = ~ 1 | id),
                 weights = varIdent(form = ~ 1 | gender),method="ML")
reduced <- gls(distance ~ -1 + gender + age, data = dat.den,
               correlation=corCompSymm(form = ~ 1 | id),
               weights = varIdent(form = ~ 1 | gender),method="ML")

anova(full,reduced)


```

## Using Wald test to compare models

```{r eval=T, echo = T, include=TRUE}
##  Note we can test the difference in slopes directly using a
##  Wald-type test
full.alt <- gls(distance ~ gender + age + gender:age, data=dat.den,
                correlation=corCompSymm(form = ~ 1 | id),
                weights = varIdent(form = ~ 1 | gender),method="ML")

summary(full.alt)


```

\newpage
# 2. Models for the Stepped Care data

### Exercise: Consider the models used here for the Stepped-Care trial and select a model that fits the data well.

**Objective** of the study was to determine if there is an intervention effect.


```{r "read stepped-care", eval=T, echo=TRUE, include=TRUE}
# Read in data 
dat.step <- read.csv("/Users/juarezce/Documents/OneDrive - The University of Colorado Denver/BIOS6643/BIOS6643_Notes/data/stepped-care-class.csv", header=TRUE)

head(dat.step, 3)
```


```{r test, eval=T, echo = F, fig.height=3.7, include=F}
  
  ggplot(dat.step, aes(x = time, y = cops, group = id)) +
  # facet_wrap(~treat) +
  geom_line(aes(color = treat), alpha = 0.2) +
  theme_classic() +
  theme(legend.position = "top") +
  ylab("Copping Skill")

```


## 



```{r  eval=T, echo = T, include=TRUE}


```



```{r  eval=T, echo = T, include=TRUE}



```

```

