---
title: "BIOS6643. L18 Models for data with many zeros"
cheauthor: "EJC"
-- date: "" 
header-includes:
   - \usepackage{amsmath}
   - \usepackage{float}
output: pdf_document
---

\newcommand{\bi}{\begin{itemize}}
\newcommand{\ei}{\end{itemize}}
\newcommand{\itt}{\item}


```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- -->


```{r installp, echo = T, eval=T, include=F}

library(glmmTMB)


```

## Salamander data
A data set containing counts of salamanders with site covariates and sampling covariates. Each of 23 sites was sampled 4 times. See Price et al. (2016, Effects of mountaintop removal mining and valley filling on the occupancy and abundance of stream salamanders. Journal of Applied Ecology)

\medskip
A data frame with 644 observations on the following 10 variables:

- *site*: name of a location where repeated samples were taken
- *mined*: factor indicating whether the site was affected by mountain top removal coal mining 
- *cover*: amount of cover objects in the stream (scaled)
- *sample*: repeated sample
- *DOP*: Days since precipitation (scaled)
- *Wtemp*: water temperature (scaled)
- *DOY*: day of year (scaled)
- *spp*: abbreviated species name, possibly also life stage
- *count*: number of salamanders observed

\medskip 

**Question of interest:**
  What is the association between mining and the number of salamanders found at sites (in terms of occuppancy and conditional abundance)?

\medskip 

```{r "data", eval=T, echo = T, include=T, out.width="90%"}
head(Salamanders, 3)
table(Salamanders$count)

```

\medskip 

## Poisson zero heavy model
We can use **glmmTMB** to fit ZI- or hurdle models for longitudinal data. Note that the zero part is not modeled longitudinally. 

```{r "ZIpoi", eval=T, echo = T, include=T, out.width="90%"}
fit.zip <- glmmTMB(count ~ mined + (1|site),
      zi=~mined,
      family=poisson, 
      data=Salamanders)

summary(fit.zip)


```
The expected mean count of salamanders is exp(1.14)=3.13 times in the sites with no mining than those sites with mining, for sites with non-zero salamanders. The odds of probability of zero structural counts for sites with no mining is exp(-1.74)=0.18 times the odds of probability of zero structural counts for sites with mining. 

\medskip 

## Comparison of ZIP versus Poisson 
```{r "ZI.comparison", eval=T, echo = T, include=T, out.width="90%"}

fit.p <- glmmTMB(count ~ mined + (1|site),
      zi=~0,
      family=poisson, data=Salamanders)

summary(fit.p)

## likelihood ratio test for whether or not the zero-inflation is needed
anova(fit.zip, fit.p)


```

\medskip 

## Hurdle Poisson model
```{r "hurdle", eval=T, echo = T, include=T, out.width="90%"}
fit.hurdle <- glmmTMB(count ~  mined + (1|site),
      zi=~ mined,
      family=truncated_poisson, 
      data=Salamanders)

## summary(fit.hurdle)

```



