---
title: "SASMarkdown"
output:
  html_document: default
  pdf_document: default
  word_document: default
---
```{r eval=FALSE, include=FALSE}
install.packages("installr")
install.packages('backports')
installr::updateR()
update.packages(ask = FALSE, checkBuilt = TRUE)
install.packages(c("xfun", "tinytex", "rmarkdown"), type = "source")
tinytex::install_tinytex()
sessionInfo()
```


```{r "function for autochange color"}
colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}

knitr::opts_chunk$set(cache.path = "cache/",
                      fig.path = "figures/")
```



# In a first code chunk, set up your SAS engine configuration.

This depends on your operating system, the version of SAS, and whether or not you have SAS installed in the default location. This example catches Windows SAS for me. In Macbook you just need to find the location of sas.exe and save the directory. 

SASmarkdown has a very short memory, it could not save the previous code chunks in memory if you are using Rmarkdown editing anything. It only works when knitr works. 
`r colorize("If you want to use SAS but make PDF or DOC files If you want to knit to PDF or DOC, you should change the engine to engine='sas'. This will NOT use ODS output from SAS.", "red")`

```{r echo=TRUE, message=FALSE, warning=FALSE}
# install.packages("SASmarkdown")
require(SASmarkdown)

# set up the options so that knit knows where you SAS executable is
# set the linesize to be easily readable on letter size paper, portrait
# and set the knir options using opts_chunk$set().

if (file.exists("C:/Program Files/SASHome/SASFoundation/9.4/sas.exe")) {
  saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
} else {
  saspath <- "sas"
}
sas_enginesetup(sashtml=sashtml)
sasopts <- "-nosplash -linesize 75"


knitr::opts_chunk$set(engine="sashtml", engine.path=saspath, 
        engine.opts=sasopts, comment=NA)

# run these commands to convince yourself that
# within this knitr session the engine changed.
knitr::opts_chunk$get()$engine
knitr::opts_chunk$get()$engine.path
knitr::opts_chunk$get()$engine.opts
```

# Example 1

use the sashlep.class data.

```{r example1, engine="sas", engine.path=saspath, engine.opts="-nosplash -ls 75"}
/* SAS code for exmple1, we are in SAS now so every comment has to change */
proc means data=sashelp.class (keep = age);
run;
```


# Example 2


```{r example2, engine="sashtmllog", engine.path="C:/Program Files/SASHome/SASFoundation/9.4/sas.exe", engine.opts="-nosplash -ls 75", comment=""}
ods graphics off;
proc corr data=sashelp.class nosimple plots=matrix;
run;
ods graphics on;
```


```{r example22, engine="sashtml", engine.path="C:/Program Files/SASHome/SASFoundation/9.4/sas.exe", engine.opts="-nosplash -ls 75", comment=""}
proc corr data=sashelp.class nosimple plots=matrix;
run;
```

For this section the code must include the "collectcode = TRUE"

```{r datastep, engine="sashtml", collectcode=TRUE, engine.path="C:/Program Files/SASHome/SASFoundation/9.4/sas.exe", engine.opts="-nosplash -ls 75"}
data class;
    set sashelp.class;
    keep age;
    run;
```

`r colorize("Without collectcode to link the code chunks, a later chunk that referenced the data in the WORK library would produce an error, but this now works. (No special option is needed for this later step.)", "red")`

```{r datastep2, engine="sashtml", engine.path="C:/Program Files/SASHome/SASFoundation/9.4/sas.exe", engine.opts="-nosplash -ls 75"}
proc means data=class;
run;
```

# datasetp3

`r colorize("This code chunk and the previous one does the same work: you can either use the (r, engine='sashtml') or use the {sashtml}", "red")`

```{sashtml datastep3, engine.path="C:/Program Files/SASHome/SASFoundation/9.4/sas.exe", engine.opts="-nosplash -ls 75", comment=""}
proc means data=class;
run;
```

You may run SAS (https://www.sas.com) code using the sas engine. You need to either make sure the SAS executable is in your environment variable PATH, or (if you do not know what PATH means) provide the full path to the SAS executable via the chunk option engine.path, e.g., engine.path = "C:\\Program Files\\SASHome\\x86\\SASFoundation\\9.3\\sas.exe". 

Remember to switch the engine back to R if you need to use R code.

```{r, engine='R'}

# this chunk has the engine set back to `R`

# run a short bit of r code
# scatterplot of cars dataset
plot(cars)

# check the current engine
knitr::opts_chunk$get()$engine

# reset the engine globally (i.e. so it will work outside of
# this chunk so you don't have to keep typing engine=`R`)
knitr::opts_chunk$set(engine="R",  engine.path=NULL, 
        engine.opts=NULL, comment=NA)

# confirm that this change was applied
knitr::opts_chunk$get()$engine
knitr::opts_chunk$get()$engine.path
knitr::opts_chunk$get()$engine.opts
```

# Try a online dataset

```{r, engine="sashtml", engine.path=saspath, engine.opts=sasopts, collectcode=T}
filename myurl url "https://www.utsc.utoronto.ca/~butler/c32/soap.txt";

proc import
  datafile=myurl
  out=soap
  dbms=dlm
  replace;
  getnames=yes;
  delimiter=" ";
```

After that, proceed as you would in the SAS IDE (or on SAS Studio online), *without* `collectcode` on the top of the code chunk:

```{r, engine="sashtml", engine.path=saspath, engine.opts=sasopts}
proc means;
  var scrap speed;
```

This works because the "collected" chunk with the `proc import` in it is added to the top of this code, so that the data set is read in again, and because it "belongs" to this chunk, the variables `scrap` and `speed` will be found. We could also run a regression in the same way:

```{r, "scatterplot", engine="sashtml", engine.path=saspath, engine.opts=sasopts}
proc reg;
  model scrap=speed;
```

(there are also supposed to be some plots which you won't see here)
and once again the reading in of the data is added behind the scenes to the top of this code. In this case, as we suspected from the scatterplot, there is a significantly positive relationship between the speed of the production line and the amount of scrap produced.

You could also have a second chunk of "collected" code. For example, you might want to run a regression, saving an output data set (say, with the residuals in it), and, later, do something with the residuals. My example below saves the leverages (along with all the original variables). The `noprint` on the first line suppresses the regression output, which we saw before and don't want to see again:

```{r, engine="sashtml", engine.path=saspath, engine.opts=sasopts, collectcode=T}
proc reg noprint;
  model scrap=speed;
  output out=saved h=leverage;
```

Because I put `collectcode=T` in *this* code chunk header, our collection of code now includes (a) reading in the data and (b) running this regression, obtaining the output data set with the leverages in it. Thus, to display the leverages in order, I now only need to do this:

```{r, engine="sashtml", engine.path=saspath, engine.opts=sasopts}
proc sort;
  by descending leverage;
proc print;
```

and everything will work. I sorted the leverages so that you can observe that the highest leverages go with the most extreme (highest or lowest) `speed` values.




